---
title: "Split Violin Plots"
author: "Tom Kelly"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{vioplox: Split Violin Plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

##Violin Plots

Therefore violin plots are a powerful tool to assist researchers to visualise data, particularly in the quality checking and exploratory parts of an analysis. Violin plots have many benefits:

- Greater flexbility for plotting variation than boxplots
- More familiarity to boxplot users than density plots
- Easier to directly compare data types than existing plots

As shown below for the `iris` dataset, violin plots show distribution information that the boxplot is unable to.

###General Set up


```{r}
library("vioplot")
```
We set up the data with two categories (Sepal Width) as follows:

```{r, message=FALSE}
data(iris)
summary(iris$Sepal.Width)
table(iris$Sepal.Width > mean(iris$Sepal.Width))
iris_large <- iris[iris$Sepal.Width > mean(iris$Sepal.Width), ]
iris_small <- iris[iris$Sepal.Width <= mean(iris$Sepal.Width), ]
```

###Boxplots

First we plot Sepal Length on its own:

```{r}
boxplot(Sepal.Length~Species, data=iris, col="grey")
```


An indirect comparision can be achieved with par:

```{r}
par(mfrow=c(2,1))
boxplot(Sepal.Length~Species, data=iris_small, col = "lightblue")
boxplot(Sepal.Length~Species, data=iris_large, col = "palevioletred")
par(mfrow=c(1,1))
```

### Violin Plots

First we plot Sepal Length on its own:

```{r}
vioplot(Sepal.Length~Species, data=iris)
```


An indirect comparision can be achieved with par:

```{r}
par(mfrow=c(2,1))
vioplot(Sepal.Length~Species, data=iris_small, col = "lightblue", plotCentre = "line")
vioplot(Sepal.Length~Species, data=iris_large, col = "palevioletred", plotCentre = "line")
par(mfrow=c(1,1))
```

### Split Violin Plots

A more direct comparision can be made with the `side` argument and `add = TRUE` on the second plot:

```{r}
vioplot(Sepal.Length~Species, data=iris_large, col = "palevioletred", plotCentre = "line", side = "right")
vioplot(Sepal.Length~Species, data=iris_small, col = "lightblue", plotCentre = "line", side = "left", add = T)
title(xlab = "Species", ylab = "Sepal Length")
legend("topleft", fill = c("lightblue", "palevioletred"), legend = c("small", "large"), title = "Sepal Width")
```

### median

The line median option is more suitable for side by side comparisions but the point option is still available also:


```{r}
vioplot(Sepal.Length~Species, data=iris_large, col = "palevioletred", plotCentre = "point", side = "right", pchMed = 21, colMed = "palevioletred4", colMed2 = "palevioletred2")
vioplot(Sepal.Length~Species, data=iris_small, col = "lightblue", plotCentre = "point", side = "left", pchMed = 21, colMed = "lightblue4", colMed2 = "lightblue2", add = T)
title(xlab = "Species", ylab = "Sepal Length")
legend("topleft", fill = c("lightblue", "palevioletred"), legend = c("small", "large"), title = "Sepal Width")
```

It may be necessary to include a `points` command to fix the median being overwritten by the following plots:

```{r}
vioplot(Sepal.Length~Species, data=iris_large, col = "palevioletred", plotCentre = "point", side = "right", pchMed = 21, colMed = "palevioletred4", colMed2 = "palevioletred2")
vioplot(Sepal.Length~Species, data=iris_small, col = "lightblue", plotCentre = "point", side = "left", pchMed = 21, colMed = "lightblue4", colMed2 = "lightblue2", add = T)
points(1:length(levels(iris$Species)), as.numeric(sapply(levels(iris$Species), function(species) median(iris_large[grep(species, iris_large$Species),]$Sepal.Length))), pch = 21, col = "palevioletred4", bg = "palevioletred2")
title(xlab = "Species", ylab = "Sepal Length")
legend("topleft", fill = c("lightblue", "palevioletred"), legend = c("small", "large"), title = "Sepal Width")
```

Similarly points could be added where a line has been used previously:


```{r}
vioplot(Sepal.Length~Species, data=iris_large, col = "palevioletred", plotCentre = "line", side = "right", pchMed = 21, colMed = "palevioletred4", colMed2 = "palevioletred2")
vioplot(Sepal.Length~Species, data=iris_small, col = "lightblue", plotCentre = "line", side = "left", pchMed = 21, colMed = "lightblue4", colMed2 = "lightblue2", add = T)
points(1:length(levels(iris$Species)), as.numeric(sapply(levels(iris$Species), function(species) median(iris_large[grep(species, iris_large$Species),]$Sepal.Length))), pch = 21, col = "palevioletred4", bg = "palevioletred2")
points(1:length(levels(iris$Species)), as.numeric(sapply(levels(iris$Species), function(species) median(iris_small[grep(species, iris_small$Species),]$Sepal.Length))), pch = 21, col = "lightblue4", bg = "lightblue2")
title(xlab = "Species", ylab = "Sepal Length")
legend("topleft", fill = c("lightblue", "palevioletred"), legend = c("small", "large"), title = "Sepal Width")
```

Here it is aesthetically pleasing and intuitive to interpret categorical differences in mean and variation in a continuous variable.


#### Sources
These extensions to `vioplot` here are based on those provided here:

* https://gist.github.com/mbjoseph/5852613

These have previously been discussed on the following sites:

* http://mbjoseph.github.io/2013/06/24/violin.html

* http://tagteam.harvard.edu/hub_feeds/1981/feed_items/209875

* https://www.r-bloggers.com/split-violin-plots/


